First paper - Draft

### 2017-06-01 paper/paper.Rnw
Moving script to paper - modelling part done

@@

<<>>=
pisa_all <- read_rds("./data/pisa_listcol.Rdata")
pisa_all2 <- pisa_all

@@

\end{document}

>>>>

  # source("./transform_data.R")

@@

<<loading_data/recoding, echo = F, cache = TRUE>>=
pisa_all <- read_rds("./data/pisa_listcol.Rdata")
pisa_all2 <- pisa_all

@@

<<escs_trend, echo = F, cache = TRUE>>=
  # Rescaled trend ESCS data to merge.
  # This only has data for seq(2000, 2012, 3) because
  # PISA 2015 has the ESCS trend variable.
  dir <- tempdir()
  file_name <- "escs_trend.zip"
  download.file("http://vs-web-fs-1.oecd.org/pisa/trend_escs_SPSS.zip",
                destfile = file.path(dir, file_name))
  unzip(file.path(dir, file_name), exdir = dir)
  escs_trend <- map(file.path(dir, list.files(dir, pattern = ".sav")), haven::read_spss)
  file.remove(file.path(dir, list.files(dir)))

  escs_trend <-
    map(escs_trend, ~ {
    mutate(.x, cnt = pisa_countrynames[cnt]) %>%
    rename(country = cnt)
  })

<<merge_escs_pisa, echo = F, cache = TRUE>>=
   # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
  # 2015) doesn't need to be merged so I exclude it with this vector
  exclude <- -6

  # Loop in parallel to the PISA data, the ESCS data and the year vector (which is seq(2012, 2015, 3))
  pisa_all2$value[exclude] <-
    pmap(list(pisa_all2$value[exclude], escs_trend, years[exclude]), function(.x, .y, .z) {

    # The escs data needs to have the key variables the same class as the
    # same data.
    escs <-
      .y %>% mutate(schoolid = as.numeric(schoolid),
                    stidstd = as.numeric(stidstd))

    # .z is the corresponding year that will be created as a column
    # And perform the same transformation of the key variables as in the ESCS data
    data_trend <-
      .x %>%
        mutate(
          year = .z,
          schoolid = as.numeric(as.character(SCHOOLID)),
          stidstd = as.numeric(as.character(STIDSTD))
          ) %>%
   left_join(escs,
              by = c("country", "schoolid", "stidstd"))

    message(paste(unique(.x$wave), "done"))

    data_trend
  })

  pisa_all2$value[[6]] <-
    pisa_all2$value[[6]] %>%
    rename(escs_trend = ESCS)
@

<<functions_for_modelling, echo = F, cache = TRUE>>=

# Function calculates the bottom 30th quantile for the bottom educated and the 70th quantile
# for the top educated. If the quantiles can't be estimated, it returns two NA's instead
quantile_missing <- function(df, weights) {

    df_lower <- filter(df, new_hisced == 1)
    df_upper <- filter(df, new_hisced == 7)

    quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
                                          weights = df_lower[[weights]],
                                          probs = c(0.30)))

    quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
                                    weights = df_upper[[weights]],
                                    probs = c(0.70)))

    if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
      return(c(NA, NA))
      } else {
     return(c(quan_lower[1], quan_upper[1]))
    }
}

# Producing the plot to get the difference between the top 30% of the high educated
# vs the bottom 30% of the low educated. This function loops through each dataset/country
# and survey reliability and estimates the difference while also calculating the
# joint standard error of the difference.

# It returns a dataframe for each survey with all countries and respective coefficients and
# standard errors.
test_diff <- function(df, reliability, test) {

    map2(df, reliability, function(.x, .y) {

      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
      weights_var <- conf$variables$weightFinal

      country_split <- split(.x, .x$country)

      country_list <- map(country_split, function(country) {
        print(unique(country$country))

        # In some countries the quan can't be estimated because of very few obs.
        # The function doesn't stop but returns two NA's.
        quan <- quantile_missing(country, weights_var)

        # It's very important to create a variable that returns the number of observations of this dummy
        # For each country. Possibly to weight by the number of observations.
        country$escs_dummy <-
          with(country, case_when(escs_trend >= quan[2] ~ 1,
                                  escs_trend <= quan[1] ~ 0))
        country
      })

      .x <-
        enframe(country_list) %>%
        unnest(value)

      .x <-
        .x %>%
        select(wave, matches(paste0("^PV.*", test, "$")), escs_dummy,
               country, one_of(weights_var),
               AGE)

      message(paste(unique(.x$wave), "data ready"))

      test_vars <- paste0("PV", seq_len(conf$parameters$PVreps), test)
      .x[test_vars] <- map(.x[test_vars], ~ ifelse(.x == 9997, NA, .x))

      # Calculate median math score of all PV's
      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)

      # Should I estimate the model separately by country?
      mod1 <- lm(dv ~ AGE,
                 weights = .x[[weights_var]],
                 data = .x,
                 na.action = "na.exclude")

      # Take residuals of model and divide by rmse. Multiply that by
      # 1 / sqrt(reliability of each survey), which is .y in the loop.
      .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)

      mod2 <-
        lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
             data = .x,
             weights = .x[[weights_var]])

      # Take the country coefficients (absolute coefficients)
      country_coef <-
        coef(mod2)$country %>%
        rownames_to_column() %>%
        gather(escs_dummy, Mean, -rowname) %>%
        mutate(escs_dummy = dplyr::recode(escs_dummy,
                                          `(Intercept)` = "0",
                                          `escs_dummy` = "1"))

      # Take the absolute country standard errors
      se <-
        se.coef(mod2)$country %>%
        as.data.frame() %>%
        rownames_to_column() %>%
        gather(escs_dummy, s.e., -rowname) %>%
        mutate(escs_dummy = dplyr::recode(escs_dummy,
                                          `(Intercept)` = "0",
                                          `escs_dummy` = "1"))

      results <-
        inner_join(country_coef, se, by = c("rowname", "escs_dummy")) %>%
        rename(country = rowname) %>%
        arrange(country, escs_dummy)

      message(paste0(unique(.x$wave), " modeling done"))
      results
    })
}
@

<<modelling, echo = F, cache = T>>=
adapted_year_data <-
    map(pisa_all2$value, ~ {
      if (unique(.x$wave) == "pisa2000") {
        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
        # is 7 and the bottom is 1
        .x <- 
          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
      } else {
        .x <-
          mutate(.x, new_hisced = as.character(high_edu_broad))
      }
      .x
})

results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
# US is missing for reading

@
\end{document}
